Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-03-28 forecastflows.tex
add some comments
continue writing...

{\bf Goal:} Increase awareness of how we can merge statistical and scientific workflows in ecology (especially forecasting) and what we would get out of it.
\vspace*{0.5cm}

@@

\vspace{0.2cm}
\noindent \textbf{Scientific practice and current workflows}

@@

% Divergences from this are common 
Divergences from this scientific method are common. One explanation is the lack of rigor, leading to questionable practices, such as retrospectively crafting hypothesis to explain the results of a model rather than testing a clear pre-defined question (data-driven analysis, Fig. 1). But the reality of ecological research is also a major driver of the divergences from the ideal scientific method. Many important questions often cannot be addressed by conducting experiments and replications, and we must often rely on existing datasets to have a large-scale and long-term perspective. However, this does not explain the persistent flaws and lack of overall coherence in ecological modeling. Trend estimation still often relies on a one-way model fitting, and most of the time is spent fitting the model to empirical data (Fig. 1). For forecasting, researchers focus on making predictions with complex models, but the steps of model building and parameterization are not very transparent and not clearly delineated (Fig. 1). The different parts of the model are often calibrated separately rather than as a whole, and some parameter values are just fixed based on experiments and expert knowledge.

% There is thus room for improvment, with a workflow that addresses theses flaws while taking account the reality of working with ecological data.

>>>>

{\noindent \bf Goal:} Increase awareness of how we can merge statistical and scientific workflows in ecology (especially forecasting) and what we would get out of it.
\vspace*{0.5cm}

@@

\vspace{0.5cm}
\noindent \textbf{Scientific method and workflows}
%  too much blah blah blah?

@@

% Divergences from this are common 
Divergences from this scientific method are common. One explanation is the lack of rigor, leading to questionable practices, such as retrospectively crafting hypothesis to explain the results of a model rather than testing a clear pre-defined question (data-driven analysis, Fig. 1). But the reality of ecological research is also a major driver of the divergences from the ideal scientific method. Many important questions often cannot be addressed by conducting experiments and replications, and we must often rely on existing datasets to have a large-scale and long-term perspective. % I feel something is missing
However, this does not explain the persistent flaws and lack of overall coherence in ecological modeling. Trend estimation still often relies on a one-way 'inference',  % need a better word? to stress out that there is no feedback, eah paper applied a slightly different model 
and most of the time is spent fitting the model to empirical data (Fig. 1). For forecasting, researchers focus on making predictions with complex models, but the steps of model building and parameterization are not very transparent and not clearly delineated (Fig. 1). The different parts of the model are often calibrated separately rather than as a whole, and some parameter values are just fixed based on experiments and expert knowledge.

% There is thus room for improvment, with a workflow that addresses theses flaws while taking account the reality of working with ecological data.
% Moving along the data-model space
% \item stressing the need to think about model before study design
% \item advancing data simulation
[...]

% More details on the workflow:  walk through the different steps
[...]

\vspace{0.5cm}
\noindent \textbf{The workflow in pratice}
% How to address current issues (two case studies)

% Trends - outline current problem
[...]

% Trends - improvements!
[...]

% Forecasting - outline current problem
Forecasting is a broad field with a diverse range of methods. Here, we focus on process-based modeling, which is often considered the gold standard in ecology and beyond. Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. Model development is the central step, typically requiring several years, yet it often remains opaque from an external perspective. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration, where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes. Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e. fix to a value found in the literature).

% Forecasting - potential improvments (do I say too much ideas)
Applying the workflow to process-based models is a key for opening the black box. It would serve as a guide through the successive steps of model development. In particular, incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and exposing strong degeneracies in the model design. This approach would force researchers to begin with a simpler version of the model, providing a clear pathway to support---or reject---the additional complexity and new parameters along the iterative development of the model.
Data fitting would no longer be just a hidden aspect of model building but a step as crucial as forecasting to gain new ecological insights. The workflow could also refocus attention on the research question, defining a clear and limited context in which the model should apply.
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own. Finally, an universal workflow provides an opportunity to merge statistical and process-based approaches, integrating mechanistic knowledge and leveraging robust statistical approaches.

% Step back
%\item we need more data, and better question (relate this to both previous study cases)
%\item where can we best reduce uncertainties through new scientific insights?
%\item machine learning! If we change nothing, what's the point of not doing ML? ML $>$ process-based without question, and ML $>$ trends without mechanisms! Where theory is lacking, or where we are far from mechanistic understanding, you might as well do ML!
%\item (ML has a way to collect and interpret large datasets...)

\vspace{0.5cm}
\noindent \textbf{Wrap up: how to make it happen?}

### 2025-03-28 forecastflows_outline.tex
continue writing...

\item More details on ideal workflow:  walk through the different steps \textbf{(Figure 2)}, and emphasizes: not rigid structure, integration of everything,  iterate the process of science, explicit effort to recognize the uncertainties, 

>>>>

\item More details on the workflow:  walk through the different steps \textbf{(Figure 2)}, and emphasizes: not rigid structure, integration of everything,  iterate the process of science, explicit effort to recognize the uncertainties, \\
forecast is just a step:  jointly model the new circumstance along with the original data
